% ============================================================================
% COPYRIGHT & IMPRINT PAGE — E-book version (smaller fonts)
% ============================================================================
% Simplified version for e-readers where screen space is at a premium.
% Uses smaller fonts to de-emphasize legal boilerplate while maintaining
% all required information.
%
% Key differences from print version:
%   - Scriptsize/footnotesize fonts throughout
%   - No CC license icons (may not render well on all e-readers)
%   - Condensed spacing
%   - Ebook ISBN listed first
% ============================================================================

\thispagestyle{empty}
\null
\vfill

{\footnotesize
\noindent\textbf{Agentic AI in Law and Finance}\\
\textbf{Navigating a New Era of Autonomous Systems}\\[0.8em]

\noindent Copyright \textcopyright\ 2025 Michael J. Bommarito II, Daniel M. Katz, and Jillian Bommarito.\\[0.8em]

% --- License (text only, no icons for ebook) ---
\noindent This work is licensed under the \textbf{Creative Commons Attribution-ShareAlike 4.0 International License} (CC~BY-SA~4.0). You may share and adapt this material for any purpose, even commercially, provided you give appropriate credit and distribute contributions under the same license.\\[0.8em]

% --- Disclaimer ---
\noindent This book is for informational purposes only and provided ``as is'' without warranty. Nothing herein constitutes legal, financial, or professional advice. The authors and publishers disclaim liability for any actions taken based on this content.\\[0.8em]

% --- Online Resources (condensed) ---
\noindent\textbf{Resources:}\\
Project: \texttt{ai4lf.com}\\
Source: \texttt{github.com/mjbommar/ai-law-finance-book}\\
License: \texttt{creativecommons.org/licenses/by-sa/4.0/}\\[0.8em]

% --- Edition and ISBNs ---
\noindent\textit{First Edition, 2025}\\[0.3em]
\noindent ISBN 979-8-9943457-0-2 (paperback)
% \noindent ISBN [ebook ISBN] (ebook)  % Uncomment when ebook ISBN is assigned
}

\vspace{1em}

% ----------------------------------------------------------------------------
% Publisher's Cataloging-in-Publication Data (condensed for ebook)
% ----------------------------------------------------------------------------
{\scriptsize\raggedright
\noindent\textbf{Publisher's Cataloging-in-Publication Data}\\[0.3em]
Names: Bommarito, Michael J., II; Katz, Daniel M.; Bommarito, Jillian, authors.\\
Title: Agentic AI in law and finance : navigating a new era of autonomous systems.\\
Description: First edition. | 2025. | Includes bibliographical references and glossary.\\
Identifiers: ISBN 979-8-9943457-0-2 (paperback)\\
Subjects: LCSH: Artificial intelligence---Law and legislation. |
Artificial intelligence---Finance. |
Intelligent agents (Computer software). |
Machine learning.\\
Classification: LCC K564.C6 | DDC 343.09/9---dc23
}

\vfill
