% ============================================================================
% Controls Catalogue — Agents Part III (Govern)
% Purpose: Practical control set mapped to COSO/COBIT/ISO 27001/SOC 2
% Label: sec:agents3-controls
%
% APEX‑2D mapping (comments only):
%   Primary: E (Evidence/Audit), A (Attribution/Provenance), P (Privilege)
%   Secondary: X (Escalation/eXit via change gates and stop/rollback)
%   Dials: Autonomy/Actuation (control strength), Jurisdiction (local controls)
% ============================================================================

\section{Controls Catalogue: From Risk to Evidence}
\label{sec:agents3-controls}

We present a practical controls set aligned to COSO, COBIT, ISO/IEC 27001, and SOC 2 Trust Services Criteria. Each control should have owners, procedures, and objective evidence. Controls are organized so they directly support Part I’s six operational properties (Iteration, Adaptation, Termination layered on Goal/Perception/Action) and the four analytical dimensions (Autonomy, Entity Frame, Goal Dynamics, Persistence).

\subsection{Control Domains}
\begin{itemize}
  \item \textbf{Data Governance (Persistence/Entity):} Classification, residency, retention, and deletion; lawful basis records tied to entity context.
  \item \textbf{Privilege and Confidentiality (Confidentiality):} Privilege‑safe memory/logging; tool‑level gating; secure enclaves.
  \item \textbf{Attribution and Verification (Perception/Action):} Source‑grounding, quote fidelity, mandatory citator checks.
  \item \textbf{Identity and Access (Autonomy):} SSO, role‑based access, least privilege, session recording for sensitive actions and higher autonomy.
  \item \textbf{Change and Model Management (Adaptation):} Versioning, release approvals, rollback; dataset lineage and drift monitoring.
  \item \textbf{Incident/BCP/DR (Termination):} Detection, response, notification, tabletop tests, recovery time objectives.
  \item \textbf{Third‑Party Risk (Entity/Provenance):} Due diligence, SOC 2/ISO evidence, subprocessor inventories, DPAs.
\end{itemize}

\subsection{Framework Alignment (Illustrative)}
\begin{center}
\begin{tabular}{p{0.35\textwidth}p{0.55\textwidth}}
\toprule
\textbf{Control} & \textbf{Alignment (examples)} \\
\midrule
Privilege‑safe memory/logging & COSO Control Activities; COBIT DSS05; ISO 27001 A.8, A.12; SOC 2 CC6/CC7/Confidentiality \\
Attribution/verification pipeline & COSO Information \,\& Communication; COBIT BAI03; ISO 27001 A.12; SOC 2 Processing Integrity \\
Change/model approvals & COSO Risk Assessment; COBIT BAI06; ISO 27001 A.12/A.14; SOC 2 CC8 \\
Third‑party due diligence & COSO Control Activities; COBIT APO10; ISO 27001 A.15; SOC 2 CC9 \\
\bottomrule
\end{tabular}
\end{center}

\begin{questionbox}[title={Evidence Checklist (per control)}]
\begin{evallist}
  \item Written procedure and control owner identified.
  \item Configuration screenshots/exports and automated test artifacts.
  \item Logs and sampled records demonstrating operation over time.
\end{evallist}
\end{questionbox}
