% ============================================================================
% Deployment Guidance (Layer 6) — Agents Part III (Govern)
% Purpose: Procurement, SLAs/KPIs, change control, data residency
% Label: sec:agents3-deploy
%
% APEX‑2D mapping (comments only):
%   Primary: X (go‑live gates, rollback, oversight modes), E (SLA/KPI evidence)
%   Secondary: P (data residency/segregation), A (verification thresholds)
%   Dials: Autonomy/Actuation (oversight choice), Jurisdiction (localization)
% ============================================================================

\section{Deployment Guidance: Procurement to Production}
\label{sec:agents3-deploy}

Layer 6 translates requirements into contracts, SLAs, and operational practices. Deployment choices explicitly depend on Part I’s \emph{Autonomy} and \emph{Actuation} risk: higher autonomy or external actuation requires stricter oversight (human‑in‑the‑loop or in‑command), narrower SLAs, and stronger rollback guarantees.

\subsection{Procurement and Due Diligence}
Request SOC 2 Type II or ISO/IEC 27001 scope with annex mappings; confirm subprocessors, data flows, and data residency. Review DPAs and IP/indemnity.

\subsection{KPIs and SLAs}
Define attribution/verification thresholds, response and resolution times, uptime, RTO/RPO, release notification windows, and rollback capabilities. Include escalation‑rate KPIs for agent workflows that \emph{adapt} goals.

\subsection{Change Control}
Require release notes, model/dataset versioning, backward compatibility guarantees, and validation results for material changes before production rollout.

\subsection{Localization and Residency}
Enforce jurisdiction‑specific processing, encryption, and access boundaries with auditable controls.

\begin{questionbox}[title={Go‑Live Checklist (Executive)}]
\begin{evallist}
  \item Contracts: DPIA/TRA completed; SLAs signed; controls mapped.
  \item Validation: Layers 4–5 passed; evidence archived.
  \item Operations: Owners, runbooks, incident and DR plans tested.
  \item Governance: Policy updates, training, and reporting cadence set.
\end{evallist}
\end{questionbox}
