% ============================================================================
% Professional and Self-Regulatory Frameworks — Agents Part III (Govern)
% Purpose: Attorney-client privilege, ABA ethics, CPA/AICPA, FINRA SRO
% Label: sec:agents3-ethics
%
% APEX‑2D mapping (comments only):
%   Primary: P (Privilege/Confidentiality), A (Attribution/Provenance), X (Escalation)
%   Secondary: E (Evidence/Audit)
%   Dials: Jurisdiction/Localization (court/local bar rules), Autonomy (supervision duties)
% ============================================================================

\section{Professional and Self‑Regulatory Frameworks}
\label{sec:agents3-ethics}

Governance must reflect professional duties that exceed generic privacy/security. Part I flags additional \emph{professional governance requirements} for high‑stakes domains—\textbf{attribution}, \textbf{provenance}, \textbf{escalation}, and \textbf{confidentiality}. This section operationalizes those requirements for legal and financial practice.

\subsection{Legal Ethics and Privilege}
\begin{itemize}
  \item \textbf{Attorney–Client Privilege/Work Product:} Boundaries for prompts, intermediate artifacts, memory, and logs; waiver risks and retention.
  \item \textbf{ABA Model Rules:} Competence (tech understanding), confidentiality, supervision of non‑lawyers/technology, communications about services.
  \item \textbf{Court Policies:} Generative use disclosures, citation verification, sealed filings handling.
\end{itemize}

\subsection{Accounting and Audit}
\begin{itemize}
  \item \textbf{CPA/AICPA Ethics and Independence:} Scope of services, documentation, evidence sufficiency; segregation between advisory and attest.
  \item \textbf{SOC Reports:} User control considerations, complementary subservice controls, Type I/II expectations for vendors.
\end{itemize}

\subsection{Financial Industry SROs}
\begin{itemize}
  \item \textbf{FINRA/Exchange Rules:} Supervision, communications, recordkeeping, surveillance; suitability and fair dealing for AI‑assisted interactions.
\end{itemize}

\begin{keybox}[title={Operationalizing Ethics and Privilege (Links to Part I)}]
\begin{itemize}
  \item \textbf{Confidentiality \& Entity Frame}: Classify data (client confidential/privileged) and gate tools accordingly; prefer on‑prem or enclave patterns for institutional frames.
  \item \textbf{Provenance \& Persistence}: Suppress retention for privileged prompts/outputs unless required; encrypt logs with access review and retention policies.
  \item \textbf{Escalation \& Autonomy}: Require human approval for external transmissions and filings; lower autonomy for high‑risk tasks.
\end{itemize}
\end{keybox}
