% ============================================================================
% Organizational Readiness — Agents Part III (Govern)
% Purpose: Policies, training, model lifecycle, reporting
% Label: sec:agents3-org
%
% APEX‑2D mapping (comments only):
%   Primary: E (reporting, accountability), X (roles/approvals)
%   Secondary: P (policy/retention), A (training on attribution norms)
%   Dials: Autonomy (segregation of duties by autonomy tier)
% ============================================================================

\section{Organizational Readiness}
\label{sec:agents3-org}

Sustainable governance requires aligned policies, training, and lifecycle processes. Reporting and training should reinforce Part I’s constructs so teams share a common vocabulary (agent vs. agentic system; autonomy; entity frames; goal dynamics; persistence).

\subsection{Policies and Standards}
Update acceptable use, data classification, secure development, and third‑party risk policies to cover agentic workflows and protocol metadata requirements.

\subsection{Training and Awareness}
Role‑specific training for attorneys, finance, risk, and engineering; just‑in‑time prompts for high‑risk actions.

\subsection{Model Lifecycle Governance}
Inventory models and datasets; track provenance; schedule periodic re‑validation; define retirement criteria.

\subsection{Reporting and Accountability}
Executive dashboards for KPIs (attribution pass rate, escalation rate, incident MTTR) and compliance attestations. Segment KPIs by autonomy level and actuation scope to target controls where risk concentrates.
