% ============================================================================
% Regulatory Landscape — Agents Part III (Govern)
% Purpose: Laws/regulations/privacy frameworks relevant to agent deployments
% Label: sec:agents3-regulatory
%
% APEX‑2D mapping (comments only):
%   Primary: A (Attribution/Provenance), E (Evidence/Audit)
%   Secondary: P (Privilege), X (Escalation/eXit)
%   Dials: Jurisdiction/Localization (strong), Autonomy/Actuation (risk tiering)
% ============================================================================

\section{Regulatory Landscape}
\label{sec:agents3-regulatory}

This section surveys statutory and regulatory frameworks that drive obligations for agentic systems in legal and financial contexts. Cite primary sources with effective dates in production drafts. Interpret each obligation through Part I’s dimensions: \emph{Autonomy} (how much discretion the system exercises), \emph{Entity Frame} (human/hybrid/machine/institutional), \emph{Goal Dynamics} (accept/adapt/negotiate), and \emph{Persistence} (statefulness across time). These characteristics determine whether a deployment is ``high‑risk'' or requires stricter controls.

\subsection{AI-Specific and Adjacent Laws}
\begin{itemize}
  \item \textbf{EU AI Act:} Risk tiers; high‑risk obligations for justice administration; conformity assessment, technical documentation, monitoring.
  \item \textbf{GDPR/UK GDPR:} Lawful basis, purpose limitation, DPIA triggers, data subject rights, cross‑border transfers, records/retention.
  \item \textbf{US Privacy:} Sectoral (HIPAA/GLBA), state privacy acts; data minimization and notice/consent; automated decision‑making disclosures where applicable.
  \item \textbf{Records/Court Rules:} Filing/format rules, service, retention, and e‑discovery obligations.
\end{itemize}

\subsection{Financial Services and Risk Guidance}
\begin{itemize}
  \item Banking regulators (FRB/OCC/FDIC) on model risk and third‑party risk; CFPB/SEC/CFTC guidance where relevant.
  \item \textbf{FINRA/Exchange SROs:} Supervisory obligations, communications/recordkeeping/surveillance, fair dealing, suitability.
\end{itemize}

\subsection{Practical Takeaways}
\begin{keybox}[title={Regulatory Must‑Haves (Mapped to Part I Concepts)}]
\begin{itemize}
  \item \textbf{Provenance \& Audit (Persistence)}: Maintain traceable provenance and audit logs for all agent actions with durable identifiers and replayability.
  \item \textbf{Impact Assessments (Autonomy/Actuation)}: Implement DPIA/TRA templates and approval gates proportional to autonomy and the system’s ability to act.
  \item \textbf{Localization (Entity Frame)}: Localize data residency and access by jurisdiction and client; encode privilege flags for human/institutional contexts.
\end{itemize}
\end{keybox}
