% ============================================================================
% CHAPTER 2: How to Design an Agent
% Architectures, Protocols, and Technical Evaluation
% ============================================================================
% This file serves as a subfile wrapper for Chapter 2 content.
% When compiled standalone, it produces a complete document.
% When included via \subfile, it integrates into the main book.
% ============================================================================

\documentclass[../../main.tex]{subfiles}

\begin{document}

% ============================================================================
% CHAPTER INTRODUCTION
% ============================================================================

\begin{highlightbox}[title={Chapter Overview}]
This chapter translates the conceptual foundations of Part I into architectural decisions. We organize agent design around \textbf{ten fundamental questions} that mirror the flow of work through any agent system:

\begin{enumerate}
  \item How does the agent know it has work? (Triggers)
  \item How does the agent understand what is being asked? (Intent)
  \item How does the agent find what it needs? (Perception)
  \item How does the agent make things happen? (Action)
  \item How does the agent remember? (Memory)
  \item How does the agent break jobs into steps? (Planning)
  \item How does the agent know when it is done? (Termination)
  \item When does the agent ask for help? (Escalation)
  \item How does the agent work with other agents? (Delegation)
  \item How do we design systems that can be governed? (Governance Surface)
\end{enumerate}

Each question maps directly to architectural components and implementation choices.
\end{highlightbox}

% ============================================================================
% CONTENT SECTIONS
% ============================================================================

\input{sections/01-introduction}
\input{sections/02-triggers}
\input{sections/03-intent}
\input{sections/04-perception}
\input{sections/05-action}
\input{sections/06-memory}
\input{sections/07-planning}
\input{sections/08-termination}
\input{sections/09-escalation}
\input{sections/10-delegation}
\input{sections/11-governance}
\input{sections/12-conclusion}

\end{document}
